---
title: "Homework 2"
author: "Hongjie Liu"
output: github_document
---


Load necessary packages for homework 2.

```{r loadpackages, message = FALSE}
library(tidyverse)
library(readxl)
```


## Problem 1

Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (`YES` vs `NO`) to a logical variable.

```{r p1_tidydata, message = FALSE}
transit_df_original = read_csv(file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")

transit_df = transit_df_original %>% 
  janitor::clean_names() %>% 
  select(line:entry, vending, ada) %>% 
  mutate(entry = recode(entry, "YES" = TRUE, "NO" = FALSE))
```

The variables that the original dataset contains are ``r names(transit_df_original)``.

First, We read the data from the csv file "NYC_Transit_Subway_Entrance_And_Exit_Data.csv". Next, we use the function `janitor::clean_names` to clean up variable names. Last, We use the functions `recode` and `mutate` to convert the entry variable from character (`YES` vs `NO`) to a logical variable.

The resulting dataset has `r nrow(transit_df)` rows and `r ncol(transit_df)` columns.

These data are still not tidy, because the values of the variables `route8` to `route11` is numeric and should be converted to character, and route number and route name are not distinct variables.

```{r p1_stnum}
stnum = transit_df %>% 
  select(station_name, line) %>% 
  distinct %>% 
  count
```

There are `r stnum` distinct stations.

```{r p1_stnum_ada}
stnum_ada = transit_df %>% 
  filter(ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct %>% 
  count
```

`r stnum_ada` stations are ADA compliant.

```{r p1_stnum_novd}
total = transit_df %>% 
  filter(vending == "NO") %>% 
  pull(entry)
  
entrynum = sum(total == TRUE)
exitnum = sum(total == FALSE)
```

The ratio of entrances to exits with no vending allow entrance is 1:`r round(exitnum/entrynum, 2)`.

Reformat data so that route number and route name are distinct variables.

```{r p1_reformat}
transit_df$route8 = as.character(transit_df$route8)
transit_df$route9 = as.character(transit_df$route9)
transit_df$route10 = as.character(transit_df$route10)
transit_df$route11 = as.character(transit_df$route11)

transit_df = transit_df %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name"
  ) %>% 
  mutate(route_number = str_replace(route_number, "route", ""))
```

```{r p1_stnum_a}
stnum_a = transit_df %>% 
  filter(route_name == "A") %>% 
  select(station_name, line) %>% 
  distinct %>% 
  count

stnum_a_ada = transit_df %>% 
  filter(route_name == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct %>% 
  count
```

`r stnum_a` stations serve the A train, of which `r stnum_a_ada` are ADA compliant.


## Problem 2

Read and clean the Mr. Trash Wheel sheet:

* specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
* use reasonable variable names
* omit rows that do not include dumpster-specific data
* round the number of sports balls to the nearest integer and converts the result to an integer variable

We assign the dataset to `mrtw_df`.

```{r p2_readtidy_mrtw}
mrtw_df = 
  read_excel(
    path = "./data/Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N549"
  ) %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(
    sports_balls = round(sports_balls),
    sports_balls = as.integer(sports_balls)
  )

mrtw_df
```

We use a similar process to import, clean, and organize the data for Professor Trash Wheel and assign it to `proftw_df`. Next, we combine this with the Mr. Trash Wheel dataset `mrtw_df` to produce a single tidy dataset `tw_df`.

```{r p2_readtidy_proftw}
proftw_df = 
  read_excel(
    path = "./data/Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M96"
  ) %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(
    trash_wheel_type = "Professor Trash Wheel",
    year = as.character(year)
  )

mrtw_df = mutate(mrtw_df, trash_wheel_type = "Mr. Trash Wheel")

tw_df = bind_rows(mrtw_df, proftw_df)

tw_df
```

Here is a description of the resulting dataset `tw_df`:

* The variables of the dataset are ``r names(tw_df)``.
* Here are some values of several key variables of the dataset: some values of the variable `weight_tons` are ``r unique(tw_df$weight_tons)[1:3]``. Some values of the variable `plastic_bottles` are ``r unique(tw_df$plastic_bottles)[1:3]``.
* The dataset has `r nrow(tw_df)` rows (number of observations) and `r ncol(tw_df)` columns.

For available data, we have:

* Professor Trash Wheel collected `r sum(pull(proftw_df, weight_tons))` tons of trash.
* Mr. Trash Wheel collected `r sum(pull(filter(mrtw_df, year == 2020), sports_balls))` sports balls in 2020.


## Problem 3

First, clean the data in pols-month.csv. Use `separate()` to break up the variable `mon` into integer variables `year`, `month`, and `day`; replace month number with month name; create a `president` variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop`; and remove the `day` variable. We assign the dataset to `pols`.

```{r p3_readtidy_pols, message = FALSE}
pols =
  read_csv("./data/pols-month.csv") %>% 
  janitor::clean_names() %>% 
  separate(mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) %>% 
  mutate(
    month = month.name[month],
    president = ifelse(prez_dem == 1, "dem", "gop")
  ) %>% 
  select(-prez_dem, -prez_gop, -day)

pols
```

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to `year` and `month`, and organize so that `year` and `month` are the leading columns. We assign the dataset to `snp`.

```{r p3_readtidy_snp, message = FALSE}
snp =
  read_csv("./data/snp.csv") %>% 
  janitor::clean_names() %>% 
  separate(date, into = c("month", "day", "year"), sep = "/", convert = TRUE) %>% 
  mutate(
    month = month.name[month],
    year = ifelse(year > 15, year + 1900, year + 2000),
    year = as.integer(year)
  ) %>% 
  select(-day) %>% 
  arrange(year, match(month, month.name)) %>% 
  select(year, month, close)

snp
```

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values. We assign the dataset to `unemployment`.

```{r p3_readtidy_unemp, message = FALSE}
unemployment =
  read_csv("./data/unemployment.csv") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment_rate"
  ) %>% 
  drop_na(unemployment_rate) %>% 
  mutate(
    month = str_to_title(month),
    month = factor(month, levels = month.abb, labels = month.name),
    month = as.character(month),
    year = as.integer(year)
  )

unemployment
```

Join the datasets by merging `snp` into `pols`, and merging `unemployment` into the result. We assign the dataset to `result`.

```{r p3_merge, message = FALSE}
result = pols %>% 
  left_join(x = ., y = snp) %>% 
  left_join(x = ., y = unemployment)

result
```

Here is a description of the three datasets `snp`, `pols`, and `employment`:

* All the three datasets have the variables `month` and `year`, which indicate the year and the month of observations respectively.
* For the dataset `snp`, the variable `president` indicates whether the president was republican or democratic; the variables `gov_gop` and `gov_dem` indicate the number of republican and democratic governors respectively; the variables `sen_gop` and `sen_dem` indicate the number of republican and democratic senators respectively; the variables `rep_gop` and `rep_dem` indicate the number of republican and democratic representatives respectively.
* For the dataset `pols`, the variable `close` indicates the closing values of the S&P stock index.
* For the dataset `unemployment`, the variable `unemployment_rate` indicates the percentage of unemployment.

Here is a description of the resulting dataset `result`:

* The dataset has `r nrow(result)` rows (number of observations) and `r ncol(result)` columns.
* The range of years is `r min(pull(result, year))`-`r max(pull(result, year))`.
* The variables of the dataset are ``r names(result)``.
